---
title: "Inference for high-throughput data"
author: "KEZHEN FEI"
date: "July 13, 2016"
output: ioslides_presentation
---


## Inference in Practice

- For several individuals in two populations
- To report which genes have different average expression levels in the two populations
- For just one gene, we could simply apply the inference techniques we learned before:
  t-test 
- Instead of one gene, we are given thousands of genes' experession data (high-throughput data). 
- What do we do? 


## p-values are random variables

- An important concept to remember
- Consider the example in which we define a p-value from a t-test with a large enough sample size to use the CLT approximation
- Our p-value is defined as the probability that a normally distributed random variable is larger, in absolute value, than the observed t-test, call it $Z$. So for a two sided test the p-value is:

$$ p = 2 \{ 1 - \Phi(\mid Z \mid)\} $$

## p-values are random variables

In R this is expressed as: 

>2*(1-pnorm(Z))

- $Z$ is a random variable 
- $\Phi$ is a deterministic function
- $p$ is also a random variable We will create a 

## Example: Monte Carlo simulation shows how the values of $p$ change

Read in the data femaleControlsPopulation.csv, then use replicate to repeatedly create p-values
```{r}
library(downloader)
filename <-"https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/femaleControlsPopulation.csv"
population <- read.csv(filename)
population <- unlist(population) # turn it into a numeric
set.seed(1)
N <- 12
B <- 10000
pvals <- replicate(B,{
  control = sample(population,N)
  treatment = sample(population,N)
  t.test(treatment,control)$p.val 
  })
hist(pvals)
```

## The histogram of p-values

```{r, echo=FALSE}
library(downloader)
filename <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/femaleControlsPopulation.csv"
#if (!file.exists(filename)) download(url,destfile=filename)
population <- read.csv(filename)
population <- unlist(population) # turn it into a numeric
set.seed(1)
N <- 12
B <- 10000
pvals <- replicate(B,{
  control = sample(population,N)
  treatment = sample(population,N)
  t.test(treatment,control)$p.val 
  })
hist(pvals)
```

##The histogram implies that the p-value is uniformly distributed

When the null hypothesis is true this is always the case

Use the CLT, the null hypothesis  $H_0$ implies that our test statistic $Z$ follows a normal distribution with mean 0 and SD 1 thus:

$$ p_a = \mbox{Pr}(Z < a \mid H_0) = \Phi(a) $$

This implies that:

$$ \begin{align} \mbox{Pr}(p < p_a) &= \mbox{Pr}[ \Phi^{-1}(p) < \Phi^{-1}(p_a) ] \ & = \mbox{Pr}(Z < a) = p_a \end{align} $$

which is the definition of a uniform distribution.

## Comparison 
In this data we have two groups denoted with 0 and 1:

```{r, echo=FALSE}
library(devtools)

install_github("genomicsclass/GSE5859Subset")

library(GSE5859Subset)
data(GSE5859Subset) ##this loads the three tables
g <- sampleInfo$group
g

dim(geneExpression)
```

## To do a comparison on a particular genen
Arbitrarily pick the one on the 25th row, we would simply compute a t-test. 
To compute a p-value, we will use the t-distribution approximation and therefore we need the population data to be approximately normal. We check this assumption with a qq-plot:

```{r}
e <- geneExpression[25,]

library(rafalib)
mypar(1,2)

qqnorm(e[g==1])
qqline(e[g==1])

qqnorm(e[g==0])
qqline(e[g==0])
```

## The qq-plot shows that the data is normally approximated. 
```{r, echo=FALSE}
e <- geneExpression[25,]

library(rafalib)
mypar(1,2)

qqnorm(e[g==1])
qqline(e[g==1])

qqnorm(e[g==0])
qqline(e[g==0])
```

## The t-test does not find this gene to be statistically significant different by the two groups:
```{r}
t.test(e[g==1],e[g==0])$p.value
```

## To answer the same question for each gene, we simply repeat the above for each gene. Here we will define our own function and use apply:
```{r}
myttest <- function(x) t.test(x[g==1],x[g==0],var.equal=TRUE)$p.value
pvals <- apply(geneExpression,1,myttest)
```

## We can now see which genes have p-values less than, say, 0.05. For example, right away we see that...
```{r}
sum(pvals<0.05)
```
So 1383 genes had p-values less than 0.05. But we have to be careful in interpreting this result because we have performed over 8,000 test.

## To demonstrate: 
If we performed the same procedure on random data, for which the null hypothesis is true for all features, we obtain the following results:

```{r}
set.seed(1)
m <- nrow(geneExpression)
n <- ncol(geneExpression)
randomData <- matrix(rnorm(n*m),m,n)
nullpvals <- apply(randomData,1,myttest)
sum(nullpvals<0.05)
```
This is to be expected: 419 is roughly 0.05*8192 and we will describe the theory that tells us why this prediction works.
More to come later in the chapter...

## Faster t-test implementation

The above implementation is very inefficient. There are several faster implementations that perform t-test for high-throughput data. We make use of a function that is not available from CRAN, but rather from the Bioconductor project.

To download and install packages from Bioconductor, we can use the install_bioc function in rafalib to install the package:
 
```{r} 
library(rafalib)
install_bioc("genefilter")
```

## Now we can show that this function "rowtest" is much faster than our code above and produce practically the same answer:
```{r}
library(genefilter)
results <- rowttests(geneExpression,factor(g))
max(abs(pvals-results$p))
```
