---
title: "GitHub Exercises"
author: "Lucas Schiffer"
date: "October 21, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We will start out today by installing GitHub and then go on to some exercises.

In this tutorial, I will have you install GitHub Desktop because it is easier to
use. Be aware that the GUI (graphical user interface) version has fewer features
than the CLI (command line interface) version.

## Installing GitHub

0. Ensure you have a GitHub username and password

1. Go to https://desktop.github.com/ and download the installer that's 
appropriate for your system.

2. Run the installer

3. Have your GitHub username and password ready

4. Open GitHub Desktop

5. When asked enter your username and password

6. You should now be ready to work

## Clone the book club "repo" (repository)

1. Open RStudio

2. File > New Project > Version Control > Git

3. Enter the book club url https://github.com/jsrodriguezl/Book_Club

4. Hit tab and the name should populate automatically

5. Hit Create Project and wait for the files to download

## Experiment with version control

1. In the top right pane of the RStudio editor you should now have a git tab, 
switch to it

2. Press the Diff button and see what happens

3. Plot a normal curve and save it as a .png file in the GitHub Session 
directory
    
    ```{r, eval=FALSE}
    x <- seq(-4, 4, length = 100)
    y <- dnorm(x)
    png(filename = "./GitHub Session/yourName.png")
    plot(x, y, main = "Standard Normal Distrubition")
    dev.off()
    ```
    
4. Now press the Diff button and see what happens

5. Check the box next to the file you've just created

6. Add a commit message something like "Lucas' plot"

7. Press the commit button and then the push button

8. Close the Diff screen and return to RStudio

9. In the top right pane and the Git tab hit pull

10. You should now be able to see the changes others have committed
