---
title: "Matrix Notation, Operations and Examples"
author: "Lavanya Kannan"
date: "April 1, 2016"
output: html_document
---
## Matrix Notations

##### 1. Create the matrix from the vector 1:1000 like this:
```{r} 
X = matrix(1:1000,100,10)
```
What is the entry in row 25, column 3?

```{r}
X[25,3]
```

##### 2. Using the function cbind, create a 10 x 5 matrix with first column x=1:10. Then add 2*x, 3*x, 4*x and 5*x to columns 2 through 5. What is the sum of the elements of the 7th row?

```{r}
x = 1:10
Y = cbind(x, 2*x, 3*x, 4*x, 5*x)
sum(Y[7,])
```

##### 3. Which of the following creates a matrix with multiples of 3 in the third column?

```{r}
A = matrix(1:60,20,3)
B = matrix(1:60,20,3,byrow=TRUE)
x=11:20; C = rbind(x,2*x,3*x)
x=1:40; D = matrix(3*x,20,2)
dim(A)
A[,3]
B[,3]
C[,3]
dim(D)
```
Answer: B

## Matrix Operations

##### 1. Suppose X is a matrix in R. Which of the following is not equivalent to X?

```{r}
X = matrix(1:15,3,5)
A = t( t(X) ) # dim (3,5)
B = X %*% matrix(1,ncol(X) ) # dim (3,5) * (5,1)
C = X*1 # dim (3,5) * (5,5)
D = X %*% diag(ncol(X)) # dim (3,5) * (5,5)

X
A
B
C
D
```
Note that all(mat1 == mat2) alone 
doesn't always do it, it only works if the matrices have equal dimensions. If they don't, it breaks the code.  That's the case of B. Another function to try is identical(mat1,mat2). However, if the matrices have different "dimnames", all(mat1 == mat2) will return TRUE, but identical(mat1, mat2) will return FALSE even if they are same. 
```{r}
identical(X,B)
Y = matrix(1:15,3,5, dimnames=list(1:3, letters[1:5]))
all(X == Y)
identical(X,Y)
```

A one line code for solving it.
```{r}
matequal <- function(x, y)
  (is.matrix(x) || is.vector(x)) && 
  (is.matrix(y) || is.vector(y)) && 
  all(dim(x) == dim(y)) && all(x == y)

matequal(X, A)   
matequal(X, B)
matequal(X, C)
matequal(X, D)
```
Answer: B


##### 2. Solve the following system of equations using R:

3a+4b−5c+d=10
2a+2b+2c−d=5
a−b+5c−5d=7
5a+d=4 

```{r}
X = rbind(c(3,4,-5,1), c(2,2,2,-1), c(1,-1,5,-5),c(5,0,0,1))
y = c(10,5,7,4)
solve(X)%*%y #equivalent to solve(X,y)
```
Answer: a = 1.2477876, b = 1.0176991, c = -0.8849558, d = -2.2389381


##### 3. What is the value in the 3rd row and the 2nd column of the matrix product of a and b?

```{r}
a <- matrix(1:12, nrow=4) # dim (4,3)
b <- matrix(1:15, nrow=3) # dim (3,5)
a
b
M <- a%*%b # dim (4,5)
M[3,2]
```

##### 3.Multiply the 3rd row of a with the 2nd column of b, using the element-wise vector multiplication with *. What is the sum of the elements in the resulting vector?

```{r}
m32 <- a[3,]*b[,2]
sum(m32) ## same as the answer of Question 2!
```

## Matrix Algebra Examples

### Suppose we are analyzing a set of 4 samples. The first two samples are from a treatment group A and the second two samples are from a treatment group B. This design can be represented with a model matrix like so:

```{r} 
X <- matrix(c(1,1,1,1,0,0,1,1),nrow=4) 
rownames(X) <- c("a","a","b","b") 
X
```


Suppose that the fitted parameters for a linear model give us:

```{r}
beta <- c(5, 2)
```

Use the matrix multiplication operator, %*%, in R to answer the following questions:

##### 1. What is the fitted value for the A samples? (The fitted Y values.)
```{r}
Y = X %*% as.matrix(beta)
Y["a",]
```

##### 2. What is the fitted value for the B samples? (The fitted Y values.)
```{r}
Y["b",]
```

##### 3. Suppose now we are comparing two treatments B and C to a control group A, each with two samples. This design can be represented with a model matrix like so:

```{r}
X <- matrix(c(1,1,1,1,1,1,0,0,1,1,0,0,0,0,0,0,1,1),nrow=6) 
rownames(X) <- c("a","a","b","b","c","c") 
X
```

Suppose that the fitted values for the linear model are given by:

```{r}
beta <- c(10,3,-3)
```
##### What is the fitted value for the B samples?
```{r}
Y = X %*% as.matrix(beta)
Y["b",]
```
##### 4. What is the fitted value for the C samples?
```{r}
Y["c",]
```