---
title: "Chapt4_lab_logistic_regression"
output: html_document
---

## 4.6.1 The Stock Market Data

Data 'Smarket' contains the percentage returns for the S&P 500 stock index over 1250 days from years 2001-2005.  For each day, the following are recorded:

'Lag1' ... 'Lag5': the percentage returns are recorded for the five previous trading days.

'Volume': number of billion shares traded on the previous day

'Today': percentage return on the current day

'Direction': indicator of whether the market was 'Up' or 'Down' on that day.


```{r}
library(ISLR)
names(Smarket)
```


```{r}
summary(Smarket)
```

```{r}
cor(Smarket[,-9])
```

```{r}
attach(Smarket)
plot(Volume)
```

## 4.6.2 Logistic Regression

Here, we fit a logistic regression model to predict 'Direction' using 'Lag1' ... 'Lag5' and 'Volume'.

Use glm(family=binomial) to run a logistic regression

```{r}
glm.fit=glm(Direction~Lag1+Lag2+Lag3+Lag4+Lag5+Volume, data=Smarket, family=binomial)
summary (glm.fit)
```

Examine the coefficients.

```{r}
coef(glm.fit)
```

Access teh p-values and other aspects of the fitted model.

```{r}
summary(glm.fit)$coef
```

Get just the p-values for the coefficients:

```{r}
summary(glm.fit)$coef[,4]
```

Predict up/down of stock market using the predict() function.

```{r}
glm.probs=predict(glm.fit,type='response')
glm.probs[1:10]
```

See that glm created a dummy variable for the Direction response variable:

```{r}
contrasts(Direction)
```

Create a vector that contains "Up" or "Down" predictions based on the logistic regression probability value.

```{r}
glm.pred=rep("Down",1250)
glm.pred[glm.probs>0.5]="Up"
```

Generate a confusion matrix to examine how many of the predictions were correctly classified.

```{r}
table(glm.pred,Direction)
(507+145)/1250
mean(glm.pred==Direction)
```

That was the training error rate! which is generally overly optimistic.  

To determine prediction accuracy, keep your training and test data separate, train on the training data, and evaluate prediction accuracy on the previously unseen test data.  Do this like so:

```{r}
train=(Year<2005)
Smarket.2005 = Smarket[!train,]
dim(Smarket.2005)
Direction.2005 = Direction[!train]
```

Now, only train on the elements seen before yr 2005, and test prediction accuracy on the remainder:

```{r}
glm.fit=glm(Direction~Lag1+Lag2+Lag3+Lag4+Lag5+Volume,
            data=Smarket, family=binomial, subset=train)
glm.probs=predict(glm.fit, Smarket.2005, type='response')
```

Determine prediction accuracy on the test data:

```{r}
glm.pred=rep("Down",252)
glm.pred[glm.probs > 0.5] = "Up"
table(glm.pred, Direction.2005)
mean(glm.pred == Direction.2005)
mean(glm.pred != Direction.2005)
```

It would appear that our prediction accuracy of 48% is worse than expected from random guessing.  

Try improving on the prediction by building a model that only includes Lag1 and Lag2, ignoring the other variables that might only serve to deteriorate prediction accuracy.

```{r}
glm.fit=glm(Direction~Lag1+Lag2, data=Smarket, family=binomial, subset=train)
glm.probs=predict(glm.fit, Smarket.2005, type='response')
glm.pred=rep("Down",252)
glm.pred[glm.probs>0.5]="Up"
table(glm.pred, Direction.2005)
mean(glm.pred == Direction.2005)
```

Note that the prediction of up (106) is correct (106/106+76) or 58% of the time.  prob(truly_up|predicted_up)  = 0.58

If we want to predict returns given specific values of Lag1 and Lag2, we would do it like so:

```{r}
predict(glm.fit, newdata=data.frame(Lag1=c(1.2,1.5), Lag2=c(1.1,-0.8)), type='response')
```

