---
title: "Babies Dataset Exercises (p. 70)"
author: |
  | Data Analysis for the Life Sciences
  | CUNY School of Public Health
  | Waldron Book Club
date: "`r format(Sys.time(), '%B %d, %Y')`"
fontsize: 12pt
output: html_document
---
  
```{r}
suppressWarnings(library(downloader))
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/babies.txt"
filename <- basename(url)
if (!file.exists(filename)) {
download(url, destfile=filename)
}
babies <- read.table("babies.txt", header=TRUE)
```

```{r}
suppressMessages(library(dplyr))
bwt.nonsmoke <- filter(babies, smoke==0) %>% select(bwt) %>% unlist
bwt.smoke <- filter(babies, smoke==1) %>% select(bwt) %>% unlist
```

```{r}
suppressWarnings(library(rafalib)) 
mean(bwt.nonsmoke)-mean(bwt.smoke)
popsd(bwt.nonsmoke)
popsd(bwt.smoke)
```

### 1. Set the seed at `1` and obtain two samples, each of size $N = 25$, from non-smoking mothers (`dat.ns`) and smoking mothers (`dat.s`). Compute a t-statistic (call it `tval`).

```{r}
N <- 25
set.seed(1)
dat.ns <- sample(bwt.nonsmoke, N)
dat.s <- sample(bwt.smoke, N)
tval <- t.test(dat.s, dat.ns)$statistic
```

### 3. Due to symmetry of the standard normal distribution, what simplified calculation do we use? 
## D

```{r}
2*pnorm(-abs(tval))
```

### 4. What quantity do we add and subtract to the estimates to obtain a 99% confidence interval? 

```{r}
(CLTQuantity <- qnorm(1-0.01/2))
```

### 5. Instead of CLT, we use the t-distribution approximation, what do we add and subtract (use `2*N-2` degrees of freedom)?

```{r}
(tQuantity <- qt(1- 0.01/2, df=(2*N-2)))
```

### 6. Why are the values from 4 and 5 so similar?  
## C
N and thus the degrees of freedom is large enough to make the normal and t-distributions very similar

### 7. Which of the sentences about a Type I error is __not__ true? 
## A
The following is another way to describe a Type I error: you decided to reject the
null hypothesis on the basis of data that was actually generated by the null hypothesis.

### 8. Set the seed at `1` and take a random sample of $N = 5$ measurements from each of the smoking and nonsmoking datasets. What is the p-value (use the `t.test` function)?

```{r}
N <- 5
set.seed(1)
dat.ns <- sample(bwt.nonsmoke, N)
dat.s <- sample(bwt.smoke, N)
t.test(dat.s, dat.ns)$p.value
```

### 9. The p-value is larger than 0.05 so using the typical cut-off, we would not reject. This is a type II error. Which of the following is not a way to decrease this type of error? As $\beta$ decreases $1-\beta$ (power) increases

```{r}
power.t.test(n = 20, delta = 1, sig.level = 0.05)
```

## C
Find a population for which the null is not true

### 10. Set the seed at 1, then use the replicate function to repeat the code used in exercise 9 10,000 times. What proportion of the time do we reject at the 0.05 level?

```{r, cache = TRUE}
B <- 10000
alpha <- 0.05
N <- 5
set.seed(1)
reject <- function(N, alpha=0.05){
    dat.ns <- sample(bwt.nonsmoke, N)
    dat.s <- sample(bwt.smoke, N)
    pval <- t.test(dat.s, dat.ns)$p.value
    pval < alpha
}
rejections <- replicate(B, reject(N))
mean(rejections)
```

### 11. Note that, not surprisingly, the power is lower than 10%. Repeat the exercise above for samples sizes of 30, 60, 90 and 120. Which of those four gives you power of about 80%?

```{r, cache = TRUE}
Ns <- seq(30, 120, 30)
set.seed(1)
powers <- sapply(Ns, function(N) {
    rejections <- replicate(B, reject(N))
    mean(rejections)
})
names(powers) <- Ns
powers
```

### 12. Repeat problem 11, but now require an $\alpha$ level of 0.01. Which of those four gives you power of about 80%?

```{r, cache = TRUE}
Ns <- seq(30, 120, 30)
set.seed(1)
powers <- sapply(Ns, function(N) {
    rejections <- replicate(B, reject(N, alpha = 0.01))
    mean(rejections)
})
names(powers) <- Ns
powers
```