---
title: "Distance Exercises"
author: "Levi Waldron"
date: "Oct 14, 2016"
output: 
  html_document: 
    number_sections: yes
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tissuesGeneExpression)
data(tissuesGeneExpression)
head(e)
head(tissue)
```

# How many biological replicates for hippocampus?

```{r}
table(tissue)
sum(tissue == "hippocampus")
```

# What is the distance between samples 3 and 45?

```{r}
dist(t(e[, c(3, 45)]))
```

# What is the distance between gene 210486_at and 200805_at?

```{r}
dist(e[c("210486_at", "200805_at"), ])
```

# If I run the command (don’t run it!):

`d = as.matrix( dist(e) )`

How many cells (number of rows times number of columns) will this matrix have?

The answer is: `r nrow(e)` x `r nrow(e)` = `r nrow(e)^2`

# Compute the distance between all pair of samples:

```{r}
d = dist( t(e) )
```

# How many distances are stored in d? Hint: What is the length of d?

```{r}
length(d)
```

# Why is the answer to exercise 5 not ncol(e)^2?

The answer is C:

```{r}
ncol(e)*(ncol(e)-1)/2
```

