---
title: "Interactions and Contrasts"
author: "Kezhen Fei"
date: "April 29, 2016"
output: ioslides_presentation 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Interactions and Contrasts
As a running example to learn about more complex linear models, we will be using a dataset which compares the different frictional coefficients on the different legs of a spider. Specifically, we will be determining whether more friction comes from a pushing or pulling motion of the leg. The original paper from which the data was provided is:

Jonas O. Wolff & Stanislav N. Gorb, Radial arrangement of Janus-like setae permits friction control in spiders, Scientific Reports, 22 January 2013.

The abstract of the paper says,

The hunting spider Cupiennius salei (Arachnida, Ctenidae) possesses hairy attachment pads (claw tufts) at its distal legs, consisting of directional branched setae… Friction of claw tufts on smooth glass was measured to reveal the functional effect of seta arrangement within the pad.

***
Figure 1 includes some pretty cool electron microscope images of the tufts. We are interested in the comparisons in Figure 4, where the pulling and pushing motions are compared for different leg pairs (for a diagram of pushing and pulling see the top of Figure 3).

We include the data in our dagdata package and can download it this way:
```{r}
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/spider_wolff_gorb_2013.csv"
filename <- "spider_wolff_gorb_2013.csv"
library(downloader)
if (!file.exists(filename)) download(url, filename)
spider <- read.csv(filename, skip=1)
```

##Initial visual inspection of the data

Each measurement comes from one of our legs while it is either pushing or pulling. So we have two variables:
```{r}
table(spider$leg,spider$type)
```

***
```{r, fig.height=6, echo=FALSE}
boxplot(spider$friction ~ spider$type * spider$leg, 
        col=c("grey90","grey40"), las=2, 
        main="Comparison of friction coefficients of different leg pairs")
```

## We see two trends:

- The pulling motion has higher friction than the pushing motion.
- The leg pairs to the back of the spider (L4 being the last) have higher pulling friction.

>- The groups have different spread around their average, we call within-group variance. This is somewhat of a problem for the kinds of linear models we will explore below, since we will be assuming that around the population average values, the errors εiεi are distributed identically, meaning the same variance within each group. The consequence of ignoring the different variances for the different groups is that comparisons between those groups with small variances will be overly “conservative” (because the overall estimate of variance is larger than an estimate for just these groups), and comparisons between those groups with large variances will be overly confident. If the spread is related to the range of friction, such that groups with large friction values also have larger spread, a possibility is to transform the data with a function such as the log or sqrt. This looks like it could be useful here, since three of the four push groups (L1, L2, L3) have the smallest friction values and also the smallest spread.

>- Some alternative tests for comparing groups without transforming the values first include: t-tests without the equal variance assumption using a “Welch” or “Satterthwaite approximation”, or the Wilcoxon rank sum test mentioned previously. However here, for simplicity of illustration, we will fit a model that assumes equal variance and shows the different kinds of linear model designs using this dataset, setting aside the issue of different within-group variances.

## A linear model with one variable


```{r}
spider.sub <- spider[spider$leg == "L1",]

fit <- lm(friction ~ type, data=spider.sub)

```

***
```{r}
summary(fit)
```

***
```{r}
(coefs <- coef(fit))
```

These two estimated coefficients are the mean of the pull observations (the referece group) and the difference between the means of the two groups. To demonstrate with R code:
```{r}
s <- split(spider.sub$friction, spider.sub$type)

mean(s[["pull"]])
```

```{r}
mean(s[["push"]]) - mean(s[["pull"]])
```

##To form the design matrix that was used inside lm:
```{r}
X <- model.matrix(~ type, data=spider.sub)
colnames(X)
```
```{r}
head(X)
```

##Design matrix (Continued)
```{r}
tail(X)
```

***
Now we’ll make a plot of the $X$ matrix by putting a black block for the 1’s and a white block for the 0’s. This plot will be more interesting for the linear models later on in this script. Along the y-axis is the sample number (the row number of the data) and along the x-axis is the column of the design matrix $X$. 

```{r, echo=FALSE}
library(rafalib)
imagemat(X, main="Model matrix for linear model with one variable")
```


***
Examining the estimated coefficients in a diagram with arrows
```{r spider_main_coef, fig.cap="^^^*The green arrow indicates the Intercept term, which goes from zero to the mean of the reference group (here the 'pull' samples). The orange arrow indicates the difference between the push group and the pull group, which is negative in this example. The circles show the individual samples, jittered horizontally to avoid overplotting.*^^^", echo=FALSE}
set.seed(1) #same jitter in stripchart
stripchart(split(spider.sub$friction, spider.sub$type), 
           vertical=TRUE, pch=1, method="jitter", las=2, xlim=c(0,3), ylim=c(0,2))
a <- -0.25
lgth <- .1
library(RColorBrewer)
cols <- brewer.pal(3,"Dark2")
abline(h=0)
arrows(1+a,0,1+a,coefs[1],lwd=3,col=cols[1],length=lgth)
abline(h=coefs[1],col=cols[1])
arrows(2+a,coefs[1],2+a,coefs[1]+coefs[2],lwd=3,col=cols[2],length=lgth)
abline(h=coefs[1]+coefs[2],col=cols[2])
legend("right",names(coefs),fill=cols,cex=.75,bg="white")
```

## A linear model with two variables
In order to model both the leg pair differences (L1, L2, L3, L4) and the push vs. pull difference, both terms need to be in the R formula. So design matrix will be formed with two variables:
```{r}
X <- model.matrix(~ type + leg, data=spider)
colnames(X)
```
```{r}
head(X)
```

***
```{r, out.height=500, fig.cap="^^^*The first column is the intercept, and so it has 1’s for all samples. The second column has 1’s for the push samples, and we can see that there are four groups of them. Finally, the third, fourth and fifth columns have 1’s for the L2, L3 and L4 samples. The L1 samples do not have a column, because L1 is the reference level for leg. Similarly, there is no pull column, because pull is the reference level for the type variable.*^^^", echo=FALSE}
imagemat(X, main="Model matrix for linear model with two factors")
```

## Model friction ~ type + leg

```{r}
fitTL <- lm(friction ~ type + leg, data=spider)

```
***
```{r, echo=FALSE}
summary(fitTL)
```