---
  title: "Handling Data in R"
  subtitle: readr, dplyr, tidyr, haven
  author: |
    | Marcel Ramos
    |  CUNY School of Public Health
    |   Hunter College
  date: "`r format(Sys.time(), '%B %d, %Y')`"
  fontsize: 10pt
  output: 
    beamer_presentation: 
      theme: "CambridgeUS"
      colortheme: "dolphin"
      fonttheme: "structureitalicserif"
      includes: 
        in_header: ~/Documents/GitHub/Book_Club/Stats_LifeSci/session1/hunter-header.txt
---

Introduction
==============================================

- Tips/Resources for learning R
- Installing packages
- Importing in data
    - readr, haven
- Manipulating data
    - dplyr, tidyr
- Mathematical Notation


Tips for learning R (general)
==============================================

* Learning R may become frustrating at times
* Learning a language
* It's a matter of practice


Useful tips for learning R (stand-alone)
==============================================
Pseudo code                                   |   Example code
--------------------------------------------  |   -------------------
install.packages(packagename)                 | install.packages("dplyr") 
library(packagename)                          | library(dplyr)
?functionname                                 | ?select
?package::functionname                        | ?dplyr::select
? 'Reserved keyword or symbol' \color{blue}{(or backticks)} | ? '%>%'
??searchforpossiblyexistingfunctionandortopic | ??simulate
help(package = "loadedpackage")               | help("dplyr")
browseVignettes("packagename")                | browseVignettes("dplyr")


Learning R via online courses
==============================================
* [\color{hunterpurple}{Coursera}](https://www.coursera.org)
* [\color{hunterpurple}{edX}](https://www.edX.org)
* [\color{hunterpurple}{RStudio tutorials}](https://www.rstudio.com/resources/training/online-learning/#R)
* [\color{hunterpurple}{Quick-R}](http://www.statmethods.net/) -
Mostly for basic and base functions
* [\color{hunterpurple}{RStudio Cheatcheets}](https://www.rstudio.com/resources/cheatsheets/)


Installing packages
==============================================

- Depends on source of package
- GitHub, CRAN, Bioconductor
- Packages
    - `utils`
    - `devtools` 
    - `BiocInstaller` 
    
    
Functions for installing packages
==============================================

```{r, eval = FALSE}
utils::install.packages("packagename")
devtools::install_github("githubuser/repository")
```


Bioconductor Packages
==============================================

- See the [\color{hunterpurple}{Bioconductor}](http://www.bioconductor.org/) site for more info

```{r, eval = FALSE}
source("https://bioconductor.org/biocLite.R")
packages <- c("packagename", "githubuser/repository", "bioconductorpackage")
BiocInstaller::biocLite()
```

- Works for CRAN, GitHub, and Bioconductor packages!

Note about installing `devtools`
==============================================

- Useful for building packages
- Download and install from GitHub
- Installation dependent on OS ([\color{hunterpurple}{Rtools}](https://cran.r-project.org/bin/windows/Rtools/) for Windows)


Functions for handling files
==============================================

- File management 

```{r, eval = FALSE}
dir()
file.path()
list.files()
untar(); unzip()
```


Using R for data importing and manipulation
==============================================

```{r, eval = FALSE}
BiocInstaller::biocLite("hadley/readr") 
browseVignettes("readr")
```

* Fast way to read in large files 

Available functions in `readr`
==============================================

- `read_delim()`

- `read_tsv()`

- `read_fwf()`

- `read_table()`

- `read_lines()`

- `read_file()`


Demo
==============================================

* session1script.Rmd

Results are in...
==============================================

1.33 Gb file

Function         | Elapsed Time (seconds)
---------------- | ------------
utils::read.csv  | ~ 137.83
readr::read_csv  | ~ 38.99


Using data.table
==============================================

- Consider using `data.table::fread()` for max performance
- data.table syntax although data.table = FALSE

```{r, eval = FALSE}
library(data.table)
?fread
```


The `haven` package
==============================================

- Great for reading in foreign data formats
- SAS, SPSS, Stata
- sas7bdat

```{r, eval = FALSE}
devtools::install_github("hadley/haven")
library(haven)
?read_sas
?read_sav
?read_dta
```

Principles of Tidy Data
==============================================
* Often said: 80% of data analysis is cleaning/munging
* Provide a standard way of organizing data\footnote{\tiny{http://vita.had.co.nz/papers/tidy-data.pdf}}
1. Each variable forms a column
2. Each observation forms a row
3. Each type of observational unit forms a table

Dataset     | Variable    | Variable
----------- | ----------- | -----------
Observation | Value       | Value
Observation | Value       | Value


Principles of Tidy Data (2)
==============================================
* Why is tidy data important? 
* Easier for the analyst and the computer to extract knowledge from a set of values
* Saves a *lot* of time

Data Munging using `tidyr`
==============================================
* `tidyr` faciliates reshaping of data
1. spread vs. gather \color{purple}{*most likely to use}\color{black}
2. extract/separate vs. unite
3. nest vs. unnest

Data Manipulation using `dplyr`
==============================================
* `dplyr` convention aims to ease cognitive burden
* Function names are easy to remember
1. select (Y)
2. mutate/transmute (add Ys / new Y)
3. filter (get Xs based on condition)
4. slice (get Xs specified)
5. summarise (reduce to single observation) 
6. arrange (re-order observations)

The `tbl_df` class and show method
==============================================
* Data frame print is messy
* `tbl_df` provides same functionality (i.e. data.frame methods work)
* Output is neat and descriptive
* See: `?tbl_df`

```{r, message = FALSE}
library(dplyr)
tbl_df(mtcars)
```

Examples of use
==============================================
* Create an example of messy data: 

```{r, echo = TRUE, message = FALSE}
library(tidyr)
data("mtcars")
mtcars <- tbl_df(mtcars)
mtcars <- select(mtcars, c(mpg:hp, wt, vs:carb))
mtcars <- unite(mtcars, cylgear, cyl, gear)
separate(mtcars, cylgear, c("cyl0", "gear0"))
```

mtcars <- select(mtcars, c(1:4, 5, 7:11))

Mutate & Transumte
==============================================
```{r, echo = TRUE, message = FALSE}
head(mutate(mtcars, displ_l = disp/61.0237), 2)
head(transmute(mtcars, disp_l = disp/61.0237),2)
```

Example with base functions
==============================================
```{r, eval = TRUE, echo = TRUE}
data("mtcars")
mtcars <- mtcars[,c("mpg", "cyl", "disp", "hp",
                    "wt", "vs", "am", "gear", "carb")]
mtcars$cylgear <- with(mtcars, paste(cyl, gear, sep = "."))
mtcars[, c("cyl1", "gear1")] <- NA
mtcars[, c("cyl1", "gear1")] <-
  t(sapply(strsplit(mtcars$cylgear, ".", fixed = TRUE), FUN = "[", c(1,2)))
head(mtcars, 3)
```

Considerations
==============================================
**Be careful of loss of information!**

* Row names were lost when converting to `table_df`
* Solution: add rownames as variable

```{r, echo = TRUE}
data(mtcars)
carrows <- rownames(mtcars)
mtcars <- tbl_df(mtcars)
mtcars <- mutate(mtcars, models = carrows)
```

Functional programming example
==============================================
```{r, eval = FALSE, echo = TRUE}
hourly_delay <- filter(
  summarise(
    group_by(
      filter(
        flights,
        !is.na(dep_delay)
      ),
      date, hour
    ), 
    delay = mean(dep_delay),
    n = n()
  ), 
  n > 10
)
```

Pipes for fluid and readable programming
==============================================
* Piping operator: %>%
* Consider the previous example with pipes:

```{r, echo = TRUE, eval = FALSE}
hourly_delay <- flights %>% 
  filter(!is.na(dep_delay)) %>%
  group_by(date, hour) %>%
  summarise(delay = mean(dep_delay), n = n()) %>%
  filter(n > 10)
```

More piping
==============================================
```{r, echo = TRUE, message = FALSE, warning = FALSE}
library(nycflights13)
flights %>% group_by(carrier) %>% 
  summarise(avg_depdelay = mean(dep_delay, na.rm = TRUE),
            count = n()) %>% left_join(airlines) %>% 
    arrange(avg_depdelay) %>% head
```

Using `separate`
==============================================
```{r , echo = TRUE, message = FALSE}
data(iris)
longdata <- gather(tbl_df(iris), key = measure, n,
  Sepal.Length:Petal.Width) %>%  separate(measure, c("type",
    "dimension"))
longdata %>% group_by(Species, type, dimension) %>%
  summarise(avg_dim = mean(n, na.rm = TRUE))
```

Piping with `tidyr`
==============================================
```{r, echo = TRUE}
library(readr)
(pew <- read_csv("./data/pew.csv"))
```

Using `gather`
==============================================

```{r, echo = TRUE}
pew %>% gather(income, n, -religion) %>% head
```

income, religion : variables to gather 
n : variable in cells 
-religion means all except religion

Using `group_by`
==============================================
```{r, echo = TRUE}
pew %>% gather(income, n, -religion) %>%
  group_by(income) %>% summarise(totals = sum(n))
```

Using `group_by` (2)
==============================================
```{r, echo = TRUE}
pew %>% gather(income, n, -religion) %>%
  group_by(religion) %>% summarise(totals = sum(n))
```

