---
title: "dplyr exercises"
author: "Levi Waldron"
date: "November 6, 2015"
output: html_document
---

# The exercises
For these exercises, we will use a new dataset related to mammalian sleep. This data is described here. Download the CSV file from this location:

We are going to read in this data, then test your knowledge of they key dplyr functions select and filter. We are also going to review two different classes: data frames and vectors.

Download the file:
```{r, cache=TRUE}
library(downloader)
download("https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/msleep_ggplot2.csv", destfile="msleep_ggplot2.csv")
```

1. Read in the msleep_ggplot2.csv file with the function read.csv and use the function class to determine what type of object is returned.

```{r}
msleep <- read.csv("msleep_ggplot2.csv")
class(msleep)
```

2. Now use the filter function to select only the primates. How many animals in the table are primates? Hint: the nrow function gives you the number of rows of a data frame or matrix.

I used `nrow()` here just to show that there are `r nrow(msleep)` rows before filtering and `r nrow(dplyr::filter(msleep, order == "Primates"))` rows after filtering:

```{r, message=FALSE}
library(dplyr)
nrow(msleep)
mprimates = filter(msleep, order == "Primates")
nrow(mprimates)
```

3. What is the class of the object you obtain after subsetting the table to only include primates?

```{r}
class(mprimates)
```

4. Now use the select function to extract the sleep (total) for the primates. What class is this object? Hint: use %>% to pipe the results of the filter function to select.

```{r}
filter(msleep, order == "Primates") %>% select(sleep_total) %>% class
```

5. Now we want to calculate the average amount of sleep for primates (the average of the numbers computed above). One challenge is that the mean function requires a vector so, if we simply apply it to the output above, we get an error. Look at the help file for unlist and use it to compute the desired average.

```{r}
filter(msleep, order == "Primates") %>% 
  select(sleep_total) %>% unlist %>% mean
```

6. For the last exercise, we could also use the dplyr summarize function. We have not introduced this function, but you can read the help file and repeat exercise 5, this time using just filter and summarize to get the answer.

